Lowestcase and Uppestcase letters: Advances in Derp Learning

### 2021-03-22 lowercase/paper/lowercase.tex
Paper almost done

@@

\title{Lowestcase and Uppestcase letters: Advances in Derp Learning}
\author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
Appears in SIGBOVIK~2021 with the
OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1 em}

@@

If so, you made use of the fact that all letters have UPPERCASE
VERSIONS (e.g.~signifying ANGER) and lowercase versions
(e.g.~signifying u dont care lol). These dimensions have other uses,
for example, it is polite to start a person's name with a capital
letter to show that you took the time to regard their humanity (as it
takes extra work to press the caps lock key, press the first letter of
their name, and then press ther caps lock key again to turn it off).
In German, Nouns start with uppercase Letters, signifying Superiority
over other grammatical Categories like Verbs and Adjectives. Lowercase
letters can be used to conserve printer ink. Actually, I'm not sure that
lowercase letters have any other uses, but let's just roll with it.

The thing is: What if I'm even MORE ANGRY THAN I WAS BEFORE? There are
some standard sorts of typographic emphasis, like I can be {\bf BOLD
  ANGRY} or \textbf{\textit{\large BIG BOLD ITALIC UNDERLINE ANGRY}}
or { \Large \textbf{\textit{\uuline{COMBINE A LOT OF THESE ANGERS}}}},
each with its own nuances, depending on the cascading style sheet or
LaTeX class file. To be even more casual than lowercase, u can learn 2
write like this, and {\scriptsize shrink away} and also \sout{cross
  out ur words in shame in advance of them even being read}, but there
are few other options for de-emphasis. Plus, when I'm FEELING PRETTY
ANGRY, TOM, how do I capitalize that already-capitalized T in order to
show the proper reverence for your humanity?

@@

% shift-reduce conflict

@@

Next, my lord, do people have bad taste! And I say this as someone who
made dozens of amateurish fonts\cite{dbzfonts} as a high school and
college student and who is contributing several new questionable
fonts\label{sec:newfonts} as a result of this paper. The database is
just filled with garbage that is unusable for this project: Fonts that
are completely illegible, fonts that are missing most of their
characters, fonts with millions of control points, Comic Sans MS,
fonts where every glyph is a drawing of a train, fonts where
everything is fine except that just the lowercase r has a width of
{\tt MAX\_INT}, and so on. So I built a UI
(Figure~\ref{fig:sortition}) for efficiently and mind-numbingly
cleaning up the database by marking fonts as broken or suitable (and
also categorizing them as serif, sans-serif, decorative, techno, etc.,
which I never used). In doing this I noticed another extremely common
problem, which was that many fonts had the same letter shapes for
uppercase and lowercase letters. This would not do for the current
application!

\begin{figure*}[ht]
\centering
  \includegraphics[width=0.9 \linewidth]{sortition}
\caption{ The interactive font data-cleaning UI. A seemingly endless
  series of fonts presents, with single keypresses putting the fonts
  into common categories such as (b)roken.
  % XXX maybe write more?
} \label{fig:sortition}
\end{figure*}

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.9 \linewidth]{casealignment}
\caption{ Example alignment to reject the font {\tt
    DrippingGooExtended}. At left, \letterform{A} (red) and
  \letterform{a} (green) rendered with the identity transform, and
  their alignment ($35\%$ difference) below. At right, the transform
  found by the black-box optimizer and the resulting alignment with
  $1.7\%$ difference. Note that the shapes are still not an exact
  match (probably errors introduced in the font export process, which
  has to round the data to integers and might apply other non-linear
  transformations like curve simplification), but these are clearly
  not a useful pair for the current
  problem.} \label{fig:casealignment}
\end{figure}

@@

We'll at least put the data in a somewhat normalized form. The neural
network will take a fixed number of inputs to a fixed number of
outputs, so a simple approach is to decide on some maximum number
control points per letter, and only try training on fonts whose
letterforms fit within that budget. Letterforms can be made of
multiple contours (e.g. a stacked \letterform{g} typically has two
holes in it, and \letterform{j} has two disjoint parts). I found that
most clean fonts had three or fewer contours, and when sorting them by
descending length, basically all of them fit within 100, 25, and 16
endpoints for the three. So, I only train on fonts where all of the
letters fit within this budget.\footnote{It would not be a good idea
  to reject only the letters that don't fit, because it might result
  in the network being trained on more \letterform{l}s (tends to be
  simple) than \letterform{g}s (tends to be complex).}

@@

\paragraph{Bonus outputs.} The output includes the predicted shape,
and also 26 individual predictors for each of the 26 letters. So a
training example is actually like \letterform{C} $\rightarrow$
\letterform{c} $[0, 0, 1, 0, 0, \ldots, 0]$, with the 1 in the third
place because C is the third letter. We don't need these outputs for
the problem itself (e.g. to lowercase new letter shapes), but there
are several ideas behind this. First, the lowercasing function we're
trying to learn does depend on the letter of the alphabet being
lowercased (in an extreme case, consider the lowercase-L
\letterform{l} and the uppercase-i \letterform{I}, which look the same
in many fonts but have different lowercase letterforms). By asking the
network to learn this as well (it is penalized when it gets the
prediction wrong), it must learn features that allow it to distinguish
different letters, and those features are available for use by outputs
we {\em do} care about. This is an easy way to coax it to learn
features that I know are meaningful without having to actually
engineer feature extractors by hand (or first train separate models,
etc.). Similarly, I could have asked it to predict whether the font
is italic, serif, the character's width, or anything else I have on
hand. Perhaps the most useful thing is that it's very clear what the
right answer is, so it gives me an easy way to see if the network is
learning anything {\em at all}. (It does.) Finally, we can do some
silly stuff with these; see Section~\ref{sec:hallucination}.

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.9 \linewidth]{trainingvector}
\caption{ Screenshot of {\tt train.exe} running on the first vector-based
  version of the problem. Shown is the \makelowercase\ model's output
  (bottom) on four shapes from four fonts (top). Some dust in between
  is the activation of the network's layers. At the very bottom, the
  26 predictions for ``what letter is this?''. The output for
  \letterform{j} is not too bad; you can see the distinct dot (a
  separate contour) and it sort of looks like a \letterform{j}. The
  \letterform{e} also has two pieces as expected but is otherwise
  garbage. The model is unsure whether the second input is an H
  or a K, and has predicted a shape sort of ambiguously between those
  two. The \lowercase{z} is also an embarrassment.
} \label{fig:trainingvector}
\end{figure}

I was so {\tiny confident} that this wouldn't work that I only trained
a \makelowercase\ model and didn't even worry about the more
complicated simultaneous training setup yet. I ran this for about
22,000 rounds, some 90 million training examples. Indeed it does not
work (Figure~\ref{fig:trainingvector}). It is not a total catastrophe,
though. We can tell from the 26 bonus outputs that the model can
clearly recognize letters (though perhaps just by memorization). Some
of the shapes it generates are along the right lines. ({\em Along the
  right lines}, get it??) I did not feel ANGRY at these results
because I expected it to not really work. Still, it ``has output'' and
so it can be used to generate a font. I made every glyph in Comic Sans
MS~\cite{comicsans} lowercase using the model (with the exception of the
\letterform{\%} character, which has too many contours---{\it five}!).
Mostly this model produces small, non-confident scrawls, like little
grains of sand, so this font is called {\bf Comic Sands}
(Figure~\ref{fig:comicsands}). The TrueType version can be downloaded
from my website and installed for your desktop publishing
needs.\footnote{Font downloads are available at
  \url{http://tom7.org/lowercase/}.}

\begin{figure*}[tp]
\centering
  \includegraphics[width=0.95 \linewidth]{comicsands}
\caption{
  Type specimen for the generated font {\bf Comic Sands}. This is
  the hateful Comic Sans MS run through an early vector-based lowercasing
  model (Section~\ref{sec:vectorversion}). At top are Comic Sans's letterforms
  \letterform{A}--\letterform{Z} run through the model and so ``made
  lowercase'' (it's obviously garbage). Next are
  \letterform{a}--\letterform{z},
  made even more lowercase. Also rubbish.
  At the bottom are the illegible pangrams
  ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
  and ``Sphinx of black quartz, judge my vow!'' Although the output
  barely resembles letters, it does have a certain dynamic
  Rorschach aesthetic, like a collection of delicate moths pinned
  to paperboard, that one could consider framing or publishing in
  the proceedings of SIGBOVIK 2021. It is certainly an improvement on
  the original font.
} \label{fig:comicsands}
\end{figure*}

@@

I also output each contour's points in a normalized order, starting
from the point closest to the origin. This removes one needless
degree of freedom.\footnote{We can see how this manifests in
  the biases on the output layer, which are a proxy for the
  ``average prediction''. In the first model, because of the
  unstructured order, these are mostly near 0.5 (center of the
  character) or 0.0 (degenerate, unused contours). In this new model,
  the distribution of biases is much more flat; it can learn that
  ``the first point tends to be near 0.25,0.25'' and ``the seventh
  point tends to be near 0.64,0.3.''}

@@

As this was training, I worked on another improvement. Ideally we
would compute the difference between the predicted shape and the
expected shape, regardless of how they're drawn. Aside from being a
bit computationally challenging, this won't really work because we
need to attribute error directly to each output in order to actually
update the model in training. I spent a valuable vacation day writing
a routine to compute the best alignment of points between the actual
and expected outputs (Figure~\ref{fig:trainingrotate}). Aside from being
harder than it looked, my alignment code ended up being pretty slow
relative to the rest of training, even worse since it ran on the CPU
instead of GPU, which reduced the training speed by $50\%$. I let it
run for 80,000 rounds, some 331 million training examples, but
eventually got bored of waiting on this approach that was slow to
train and seemed like a complicated version of an bad, oversimplified
approach. So, I control-C'd that thing and threw this whole endeavour
in the trash! But I must have confused the Recycle~Bin icon with the
fairly complicated export-to-TrueType Font process that I built,
because I ran the model on the venerable Futura~\cite{futura} font and
generated {\bf Futurda} (Figure~\ref{fig:futurda}).
%
% \begin{figure}[ht]
% \centering
%   \includegraphics[width=0.5 \linewidth]{rotatej}
% \caption{ Example alignment (from debug UI) of predicted shape (blue)
%   to expected shape (green). We require each point to be mapped (red)
%   to a point from the expected contour in a monotonic order (but
%   several can be mapped to the same one), so that we can attribute
%   error to each point. I hope it was fun to write this code because
%   it was too slow to do pretty much anything but generate this figure!
% } \label{fig:rotatej}
% \end{figure}
%

@@

% the sdf problem

@@

Don't give up! The fixed-size input/output of neural networks is
better suited to something like an array of pixels, and fonts can of
course be represented this way as well. To stay in the realm of what
my desktop computer with a single GeForce 1080 can do, I wanted to
keep the number of inputs and outputs pretty small. There's already an
excellent technique for representing font data as compact bitmaps,
which comes from computer graphics, called Signed Distance Fields
(SDFs)~\cite{green2007improved}. In a standard rasterization of a
font, each pixel of a bitmap contains 1 or 0, or perhaps an
anti-aliased value in-between. In an SDF, the bitmap instead contains
the distance to the nearest edge, and is signed (e.g. values inside
the shape are $> 0$, value outside are $< 0$). Actually in
practice we offset and saturate the values so that they are all in
$[0,1]$ (or bytes in $[0,255]$), with some nonzero ``on-edge value''
(say, 0.5) standing for ``distance 0''. In order to display the font
at the size of your choice, you then resize the SDF image with bilinear
interpolation, and then simply threshold the image. This works
surprisingly well (Figure~\ref{fig:sdf}).

\begin{figure}[ht]
\centering
  \includegraphics[width=0.95 \linewidth]{sdf-figure}
\caption{
  The signed distance function representation of a letterform.
  At the very left, a $36\times 36$ pixel rasterization of
  the character without anti-aliasing, for comparison. Any
  scaling of this will have chunky pixel artifacts. Next,
  a $36\times 36$ pixel SDF of same. Third, simply scaling
  that $36\times 36$ image to $180 \times 180$ pixels with
  bilinear sampling. Finally, that image thresholded to
  produce a $180 \times 180$ pixel rasterization, which is
  far superior despite being derived from a $36 \times 36$ pixel
  bitmap. Typically this process is performed at an even higher
  scale and then downsampled to produce an anti-aliased image.
} \label{fig:sdf}
\end{figure}

@@

I computed some stats on the font database, and determined the
following parameters for the fixed-size SDFs we train on. The images
are $36 \times 36$ pixels. The character box is placed such that there
are $2$ pixels of top padding, and $9$ pixels of left and bottom
padding. The character box is only ``nominal'' in the sense that the
font's contours can exceed its bounds, and this is completely normal
for a letter like \letterform{j} (which goes below the baseline and
often hangs to the left of the origin as well). I used an ``on-edge
value'' of 0.862 (because much more of the SDF is outside the letter
than inside) and the distance is scaled as 0.059 units per pixel (so
that pixels on the outer edge often have non-zero values). Compared to
the first version, I was somewhat more permissive in what fonts I
trained on, since there was no inherent limit to the number of
contours or their complexity. I did exclude fonts whose rasterizations
exceeded the bounds of the SDF, which is possible (very wide
\letterform{W} or low-descending \letterform{j} perhaps) but rare.

@@

I also adapted my weird home-grown software to train the
\makeuppercase\ and \makelowercase\ models simultanously. Two models
fit easily in GPU memory, with plenty of space for a stream of
training data (one training instance is only about 10kb). The only
challenging thing is arranging for them to feed each other generated
``inversion'' examples (Figure~\ref{fig:cotraining}), but this is just
a matter of programming, thank God. I should remember to do projects
that are mostly a matter of programming. Each round, $25\%$ of the
batch consists of inverted examples from the symmetric model's output
from a recent round. Training happens asynchronously, but I make sure
that one model is not allowed to get more than 2 rounds ahead of the
other, because I want this feedback loop to be somewhat tight.

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.99 \linewidth]{sdfmergederror}
\caption{ The training error for the SDF models. The red curve is the
  \makeuppercase\ model, which generally has a higher error rate
  (perhaps simply because uppercase letters usually have more pixels
  set) and blue is \makelowercase. The first few rounds have error
  that's off the charts, well above 100. The most dramatic event is
  around round 200,000, where I reduced the weight decay factor to
  $0.999995$ (from $0.9995$). I guess you just need more nines to be
  more reliable. There are some other visible peaks, which occur when
  I do things like remove nodes with very low weights or which are
  almost never activated (Section~\ref{sec:fiddly}). These momentarily
  increase error but it is easily fine-tuned away (e.g. by learning
  new biases). The peak at around 1.4M rounds is when I added a new
  layer to the end of the model, which does seem to create a new
  training regime (clear downward slope now); but this also
  significantly increases the training cost per round. Even after
  2,000,000 rounds, the network is still apparently improving, but at
  a speed of about 1 pixel loss per several weeks. Eventually the
  extremely strict SIGBOVIK deadlines mean you just have to call it done.
} \label{fig:sdfmergederror}
\end{figure}

\begin{figure}[ht]
\centering
  \includegraphics[width=0.99 \linewidth]{lastlayer}
  \caption{ The bottom-right corner of the weight matrix for the final
    layer of the network. This layer was added to the network after
    1.36 million rounds, initially as the identity matrix, and so can
    be thought of partly as a correction of the network's output prior
    to that round (though the remainder of the network continues to
    evolve). The x-axis is the output nodes, and the y-axis is the
    nodes of the previous layer. Note for example that the last 26
    columns look pretty different; these are the predictors for the 26
    letters, which occur in the output after the SDF pixels. Green
    means positive and red means negative, so if you are looking at
    this in a black-and-white printout, that may explain your current
    confusion. The exact diagonal is a strong green, close to 1.0,
    although over time these weights do diverge from the identity
    somewhat. In the bottom-right corner of size $26^2$, we are
    looking at how the 26 letter predictors are derived from the
    previous layer's predictions. We see that most letters are
    negatively correlated (makes sense; only one will ever be 1.0)
    although there are some oddities (probably because it found some
    other, better correlates). Nodes on this new layer have dense
    references to these 26 predictions on the previous; this means
    that the bottom row kind of represents biases for each of the 26
    letters (what does an average 'e' look like?). I also included a
    dense region above that, but this appears to have simply evolved
    the same way as other $36^2$ chunks have (the rest are sparse).
    These chunks have a large amount of spatial similarity (suggesting
    that the sparse sampling would be adequate), with a meaning like
    ``if this area of the image is bright, then this pixel should
    be less bright.'' It is interesting that the pixels immediately
    next to the diagonal are almost always strongly negative
    (sharpening operation). ``Thank you for attending my TED talk.''
    ---\,Figure~\ref{fig:lastlayer}}
  \label{fig:lastlayer}
\end{figure}

@@

Making the GPU code faster has really high value (could save weeks of
waiting). Since I am using OpenCL (whoa, yeah, stop me right there, I
know) I found a good trick was to generate different OpenCL code with
constants baked for each layer (for example their size and
\verb+indices_per_node+); this allows the cmopiler to use faster
tricks in inner loops for e.g.~multiplication by a compile-time constant
instead of depending on an argument or data. I have different routines
for sparse and dense layers. It might even make sense to recompile the
kernels for other parameters that change over the lifetime of
training, like the learning rate. The {\tt fma} instruction (so named
for the physical law $F=MA$) is a bit faster than
\verb|potential += w * v|, and I guess the compiler can't do this
itself because of IEEE horrors. But like, who cares? In my opinion you
should be able to put it in ``fast machine learning mode'' where it
readily makes precision errors, with a command-line option like
\verb+--fml+. With all the tweaking, the easiest win was to use the
{\tt restrict} keyword on arguments to tell the compiler that the
input cannot alias the output, for example; this presumably helps
it schedule instructions better.

@@

@@

All right, but why stop there??

>>>>

\newcommand\upsigbovik{
  \includegraphics[height=1em]{upsigbovik}
}

@@

\title{Lowestcase and Uppestcase letters: Advances in Derp Learning}
\author{Dr.~Tom~Murphy~VII~Ph.D.}\thanks{
Copyright \copyright\ 2021 the Regents of the Wikiplia Foundation.
Appears in \upsigbovik~2021 with the
OS2TypoLinegap of the Association for Computational Heresy; {\em IEEEEEE!}
press, Verlag-Verlag volume no.~0x40-2A. 1 em}

@@

If so, you made use of the fact that all letters have UPPERCASE
VERSIONS (e.g.~signifying ANGER) and lowercase versions
(e.g.~signifying u dont care lol). These dimensions have other uses,
for example, it is polite to start a person's name with a capital
letter to show that you took the time to regard their humanity (as it
takes extra work to press the caps lock key, press the first letter of
their name, and then press the caps lock key again to turn it off).
In German, Nouns start with uppercase Letters, signifying Superiority
over other grammatical Categories like Verbs and Adjectives. Lowercase
letters can be used to conserve printer ink. Actually, I'm not sure that
lowercase letters have any other uses, but let's just roll with it.

There's nothing wrong with this (despite the classical advice to use
shift to {\em reduce} conflict~\cite{aho1986compilers}). But the thing is:
What if I'm even MORE ANGRY THAN I WAS BEFORE? There are some standard
sorts of typographic emphasis, like I can be {\bf BOLD ANGRY} or
\textbf{\textit{\large BIG BOLD ITALIC UNDERLINE ANGRY}} or { \Large
  \textbf{\textit{\uuline{COMBINE A LOT OF THESE ANGERS}}}}, each with
its own nuances, depending on the cascading style sheet or LaTeX class
file. To be even more casual than lowercase, u can learn 2 write like
this, and {\scriptsize shrink away} and also \sout{cross out ur words
  in shame in advance of them even being read}, but there are few
other options for de-emphasis. Plus, when I'm FEELING PRETTY ANGRY,
TOM, how do I capitalize that already-capitalized T in order to show
the proper reverence for your humanity?

@@

@@

Next, my lord, do people have bad taste! And I say this as someone who
made dozens of amateurish fonts~\cite{dbzfonts} as a high school and
college student and who is contributing several new questionable
fonts\label{sec:newfonts} as a result of this paper. The database is
just filled with garbage that is unusable for this project: Fonts that
are completely illegible, fonts that are missing most of their
characters, fonts with millions of control points, Comic Sans MS,
fonts where every glyph is a drawing of a train, fonts where
everything is fine except that just the lowercase r has a width of
{\tt MAX\_INT}, and so on. So I built a UI
(Figure~\ref{fig:sortition}) for efficiently and mind-numbingly
cleaning up the database by marking fonts as broken or suitable (and
also categorizing them as serif, sans-serif, decorative, techno, etc.,
which classifications I never used). In doing this I noticed another
extremely common problem, which was that many fonts had the same
letter shapes for uppercase and lowercase letters. This would not do
for the current application!

\begin{figure}[tp]
\centering
  \includegraphics[width=0.98 \linewidth]{sortition}
\caption{ The interactive font data-cleaning UI. A seemingly endless
  series of fonts presents, with single keypresses putting the fonts
  into common categories such as (b)roken.
  % XXX maybe write more?
} \label{fig:sortition}
\end{figure}

@@

\begin{figure}[tp]
\centering
  \includegraphics[width=0.9 \linewidth]{casealignment}
\caption{ Example alignment to reject the font {\tt
    DrippingGooExtended}. At left, \letterform{A} (red) and
  \letterform{a} (green) rendered with the identity transform, and
  their alignment ($35\%$ difference) below. At right, the transform
  found by the black-box optimizer and the resulting alignment with
  $1.7\%$ difference. Note that the shapes are still not an exact
  match (probably noise introduced in the font export process, which
  has to round the data to integers and might apply other non-linear
  transformations like curve simplification), but these are clearly
  not a useful pair for the current
  problem.} \label{fig:casealignment}
\end{figure}

@@

We'll at least put the data in a somewhat normalized form. The neural
network will take a fixed number of inputs to a fixed number of
outputs, so a simple approach is to decide on some maximum number
control points per letter, and only try training on fonts whose
letterforms fit within that budget. Letterforms can be made of
multiple contours (e.g.~a stacked \letterform{g} typically has two
holes in it, and \letterform{j} has two disjoint parts). I found that
most clean fonts had three or fewer contours, and when sorting them by
descending length, basically all of them fit within 100, 25, and 16
endpoints for the three. So, I only train on fonts where all of the
letters fit within this budget.\footnote{It would not be a good idea
  to reject only the letters that don't fit, because it might result
  in the network being trained on more \letterform{l}s (tends to be
  simple) than \letterform{g}s (tends to be complex).}

@@

\paragraph{Bonus outputs.} The output includes the predicted shape,
and also 26 individual predictors for each of the 26 letters. So a
training example is actually like \letterform{C} $\rightarrow$
\letterform{c} $[0, 0, 1, 0, 0, \ldots, 0]$, with the 1 in the third
place because C is the third letter. We don't need these outputs for
the problem itself (e.g.~to lowercase new letter shapes), but there
are several ideas behind this. First, the lowercasing function we're
trying to learn does depend on the letter of the alphabet being
lowercased (in an extreme case, consider the lowercase-L
\letterform{l} and the uppercase-i \letterform{I}, which look the same
in many fonts but have different lowercase letterforms). By asking the
network to learn this as well (it is penalized when it gets the
prediction wrong), it must learn features that allow it to distinguish
different letters, and those features are available for use by outputs
we {\em do} care about. This is an easy way to coax it to learn
features that I know are meaningful without having to actually
engineer feature extractors by hand (or first train separate models,
etc.). Similarly, I could have asked it to predict whether the font
is italic, serif, the character's width, or anything else I have on
hand. Perhaps the most useful thing is that it's very clear what the
right answer is, so it gives me an easy way to see if the network is
learning anything {\em at all}. (It does.) Finally, we can do some
silly stuff with these; see Section~\ref{sec:hallucination}.

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.9 \linewidth]{trainingvector}
\caption{ Screenshot of {\tt train.exe} running on the first vector-based
  version of the problem. Shown is the \makelowercase\ model's output
  (bottom) on four shapes from four fonts (top). Some dust in between
  is the activation of the network's layers. At the very bottom, the
  26 predictions for ``what letter is this?''. The output for
  \letterform{j} is not too bad; you can see the distinct dot (a
  separate contour) and it sort of looks like a \letterform{j}. The
  \letterform{e} also has two pieces as expected but is otherwise
  garbage. The model is unsure whether the second input is an H
  or a K, and has predicted a shape sort of ambiguously between those
  two. The \letterform{z} is also an embarrassment.
} \label{fig:trainingvector}
\end{figure}

I was so {\tiny confident} that this wouldn't work that I only trained
a \makelowercase\ model and didn't even worry about the more
complicated simultaneous training setup yet. I ran this for about
22,000 rounds, some 90 million training examples. Indeed it does not
work (Figure~\ref{fig:trainingvector}). It is not a total catastrophe,
though. We can tell from the 26 bonus outputs that the model can
clearly recognize letters (though perhaps just by memorization). Some
of the shapes it generates are along the right lines. ({\em Along the
  right lines}, get it??) I did not feel ANGRY at these results
because I expected it to not really work. Still, it ``has output'' and
so it can be used to generate a font. I made every glyph in Comic Sans
MS~\cite{comicsans} lowercase using the model (with the exception of the
\letterform{{\sf \%}} character, which has too many contours---{\it five}!).
Mostly this model produces small, non-confident scrawls, like little
grains of sand, so this font is called {\bf Comic Sands}
(Figure~\ref{fig:comicsands}). The TrueType version can be downloaded
from my website and installed for your desktop publishing
needs.\footnote{Font downloads are available at
  \url{http://tom7.org/lowercase/}.}

\begin{figure*}[tp]
\centering
  \includegraphics[width=0.95 \linewidth]{comicsands}
\caption{
  Type specimen for the generated font {\bf Comic Sands}. This is
  the hateful Comic Sans MS run through an early vector-based lowercasing
  model (Section~\ref{sec:vectorversion}). At top are Comic Sans's letterforms
  \letterform{A}--\letterform{Z} run through the model and so ``made
  lowercase'' (it's obviously garbage). Next are
  \letterform{a}--\letterform{z},
  made even more lowercase. Also rubbish.
  At the bottom are the illegible pangrams
  ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
  and ``Sphinx of black quartz, judge my vow!'' Although the output
  barely resembles letters, it does have a certain wispy
  Rorschach aesthetic, like a collection of delicate moths pinned
  to paperboard, that one could consider framing or publishing in
  the proceedings of SIGBOVIK 2021. It is certainly an improvement on
  the original font.
} \label{fig:comicsands}
\end{figure*}

@@

I also output each contour's points in a normalized order, starting
from the point closest to the origin. This removes one needless degree
of freedom.\footnote{We can see (well, it's not pictured since I have
  far exceeded a reasonable number of figures in this paper, but {\em
    I} can see) how this manifests in the biases on the output layer,
  which are a proxy for the ``average prediction''. In the first
  model, because of the unstructured order, these are mostly near 0.5
  (center of the character) or 0.0 (degenerate, unused contours). In
  this new model, the distribution of biases is much more flat; it can
  learn that ``the first point tends to be near 0.25,0.25'' and ``the
  seventh point tends to be near 0.64,0.3.''}

@@

As this was training, I worked on another improvement. Ideally we
would compute the difference between the predicted shape and the
expected shape, regardless of how they're drawn. Aside from being a
bit computationally challenging, this won't really work because we
need to attribute error directly to each output in order to actually
update the model in training. I spent a valuable vacation day writing
a routine to compute the best alignment of points between the actual
and expected outputs (Figure~\ref{fig:trainingrotate}). Aside from being
harder than it looked, my alignment code ended up being pretty slow
relative to the rest of training, even worse since it ran on the CPU
instead of GPU, which reduced the training speed by $50\%$. I let it
run for 80,000 rounds, some 331 million training examples, but
eventually got bored of waiting on this approach that was slow to
train and seemed like a complicated version of an bad, oversimplified
approach. So, I control-C'd that thing and threw this whole endeavour
in the trash! But I must have confused the Recycle~Bin icon with the
fairly complicated export-to-TrueType Font process that I built,
because I ran the model on the venerable Futura~\cite{futura} font and
generated {\bf Futurda} (Figure~\ref{fig:futurda}).

@@

@@

Don't give up! The fixed-size input/output of neural networks is
better suited to something like an array of pixels, and fonts can of
course be represented this way as well. To stay in the realm of what
my desktop computer with a single GeForce 1080 can do, I wanted to
keep the number of inputs and outputs pretty small. There's already an
excellent technique for representing font data as compact bitmaps,
which comes from computer graphics, called Signed Distance Fields
(SDFs)~\cite{green2007improved}. In a standard rasterization of a
font, each pixel of a bitmap contains 1 or 0, or perhaps an
anti-aliased value in-between. In an SDF, the bitmap instead contains
the distance to the nearest edge, and is signed (e.g.~values inside
the shape are $> 0$, value outside are $< 0$). Actually in
practice we offset and saturate the values so that they are all in
$[0,1]$ (or bytes in $[0,255]$), with some nonzero ``on-edge value''
(say, 0.5) standing for ``distance 0''. In order to display the font
at the size of your choice, you then resize the SDF image with bilinear
interpolation, and then simply threshold the image. This works
surprisingly well (Figure~\ref{fig:sdf}).

\begin{figure}[ht]
\centering
  \includegraphics[width=0.95 \linewidth]{sdf-figure}
\caption{
  The signed distance function representation of a letterform.
  At the very left, a $36\times 36$ pixel rasterization of
  the character without anti-aliasing, for comparison. Any
  scaling of this will have chunky pixel artifacts. Next,
  a $36\times 36$ pixel SDF of same. Third, simply scaling
  that $36\times 36$ image to $180 \times 180$ pixels with
  bilinear sampling. Finally, that image thresholded to
  produce a $180 \times 180$ pixel rasterization, which is
  far superior despite being derived from a $36 \times 36$ pixel
  image. Typically this process is performed at an even higher
  scale and then downsampled to produce an anti-aliased image.
} \label{fig:sdf}
\end{figure}

@@

I computed some stats on the font database, and determined the
following parameters for the fixed-size SDFs we train on. The images
are $36 \times 36$ pixels. The character box is placed such that there
are $2$ pixels of top padding, and $9$ pixels of left and bottom
padding. The character box is only ``nominal'' in the sense that the
font's contours can exceed its bounds, and this is completely normal
for a letter like \letterform{j} (which goes below the baseline and
often hangs to the left of the origin as well). I used an ``on-edge
value'' of 0.862 (because much more of the SDF is outside the letter
than inside) and the distance is scaled as 0.059 units per pixel
(chosen so that pixels on the outer edge often have non-zero values).
Compared to the first version, I was somewhat more permissive in what
fonts I trained on, since there was no inherent limit to the number of
contours or their complexity. I did exclude fonts whose rasterizations
exceeded the bounds of the SDF, which is possible (very wide
\letterform{W} or low-descending \letterform{j} perhaps) but rare.

@@

I also adapted my weird home-grown software to train the
\makeuppercase\ and \makelowercase\ models simultanously. Two models
fit easily in GPU memory, with plenty of space for a stream of
training data (one training instance is only about 10kb). The only
challenging thing is arranging for them to feed each other generated
``inversion'' examples (Figure~\ref{fig:cotraining}), but this is just
a matter of programming, thank god. I should remember to do projects
that are mostly a matter of programming. Each round, $25\%$ of the
batch consists of inverted examples from the symmetric model's output
from a recent round. Training happens asynchronously, but I make sure
that one model is not allowed to get more than 2 rounds ahead of the
other, because I want this feedback loop to be somewhat tight.

@@

\begin{figure}[ht]
\centering
  \includegraphics[width=0.99 \linewidth]{sdfmergederror}
\caption{ The training error for the SDF models. The red curve is the
  \makeuppercase\ model, which generally has a higher error rate
  (perhaps simply because uppercase letters usually have more pixels
  set) and blue is \makelowercase. The first few rounds have error
  that's off the charts, well above 100. The most dramatic event is
  around round 200,000, where I reduced the weight decay factor to
  $0.999995$ (from $0.9995$). I guess you just need more nines to be
  more reliable. There are some other visible peaks, which occur when
  I do things like remove nodes with very low weights or which are
  almost never activated (Section~\ref{sec:fiddly}). These momentarily
  increase error but it is easily fine-tuned away (e.g.~by learning
  new biases). The peak at around 1.4M rounds is when I added a new
  layer to the end of the model, which does seem to create a new
  training regime (clear downward slope now); but this also
  significantly increases the training cost per round. Even after
  2,000,000 rounds, the network is still apparently improving, but at
  a speed of about 1 pixel loss per several weeks. Eventually the
  extremely strict SIGBOVIK deadlines mean you just have to call it done.
} \label{fig:sdfmergederror}
\end{figure}

\begin{figure}[ht!b]
\centering
  \includegraphics[width=0.99 \linewidth]{lastlayer}
  \caption{ The bottom-right corner of the weight matrix for the final
    layer of the network. This layer was added to the network after
    1.36 million rounds, initially as the identity matrix, and so can
    be thought of partly as a correction of the network's output prior
    to that round (though the remainder of the network continues to
    evolve). The x-axis is the output nodes, and the y-axis is the
    nodes of the previous layer. Note for example that the last 26
    columns look pretty different; these are the predictors for the 26
    letters, which occur in the output after the SDF pixels. Green
    means positive and red means negative, so if you are looking at
    this in a black-and-white printout, that may explain your current
    confusion. The exact diagonal is a strong green, close to 1.0,
    although over time these weights do diverge from the identity
    somewhat. In the bottom-right corner of size $26^2$, we are
    looking at how the 26 letter predictors are derived from the
    previous layer's predictions. We see that most letters are
    negatively correlated (makes sense; only one will ever be 1.0)
    although there are some oddities (probably because it found some
    other, better correlates). All nodes on this new layer have dense
    references to these 26 predictions on the previous; this means
    that the bottom 26 rows kind of represent biases for each of the 26
    letters (what does an average 'e' look like?). I also included a
    dense region above that, but this appears to have simply evolved
    the same way as other $36^2$ chunks have (the rest are sparse).
    These chunks have a large amount of spatial similarity (suggesting
    that the sparse sampling would be adequate), with a meaning like
    ``if this area of the image is bright, then this pixel should
    be less bright.'' It is interesting that the pixels immediately
    next to the diagonal are almost always strongly negative
    (sharpening operation). ``Thank you for attending my TED talk.''
    ---\,Figure~\ref{fig:lastlayer}}
  \label{fig:lastlayer}
\end{figure}

@@

Making the GPU code faster has really high value (could save weeks of
waiting). Since I am using OpenCL (whoa, yeah, stop me right there, I
know) I found a good technique was to generate different OpenCL code with
constants baked for each layer (for example their size and
\verb+indices_per_node+); this allows the compiler to use faster
tricks in inner loops for e.g.~multiplication by a compile-time constant
instead of depending on an argument or data. I have different routines
for sparse and dense layers. It might even make sense to recompile the
kernels for other parameters that change over the lifetime of
training, like the learning rate. The {\tt fma} instruction (so named
for the physical law $F=MA$) is a bit faster than
\verb|potential += w * v|, and I guess the compiler can't do this
itself because of IEEE horrors. But like, who cares? In my opinion you
should be able to put it in ``fast machine learning mode'' where it
readily makes precision errors, with a command-line option like
\verb+--fml+. With all the tweaking, the easiest win was to use the
{\tt restrict} keyword on arguments to tell the compiler that the
input cannot alias the output, for example; this presumably helps
it schedule instructions better.

@@

Ultimately, each of the two models was trained for over 2 million
rounds, which corresponds to 510 million training examples. Each
model is about 24 megabytes.

@@

Oh, you don't like letters that look {\em bad}? Instead you want
letters that look {\em good}? How about {\em best}?
